A Decision Procedure for String Constraints with String-Integer Conversion and Flat Regular Constraints

### 2021-05-10 FMCAD2021/4_dec_proc.tex
update sec 4

%\subsection{Decidability of $\paexp$} \label{sec:3.2}

\begin{definition}
    For a strictly increasing function $f:\mathbb{N}\mapsto \mathbb{N}$, 
    $f$ is said to be compatible with addition 
    if for every $m\in \mathbb{N}^+$,  $f$ modulo $m$ is periodic, 
    and for any term $A(x)=\sum_{1\le i\le n} a_i f(x+b_i)$, 
    where $n \in \mathbb{N}^+, a_i ,b_i \in \mathbb{Z}$,
    one of the following holds:
    \begin{itemize}
        \item $A(x)$ is bounded; 
        \item there exists a constant $\Delta_A\in \mathbb{N}^+$ such that 
        $\forall x. A(x+\Delta_A)\ge f(x)$;
        \item there exists a constant $\Delta_A\in \mathbb{N}^+$ such that 
        $\forall x. -A(x+\Delta_A)\ge f(x)$.
    \end{itemize}
\end{definition}

Semenov proved that for any function $f$ that is compatible with addition, 
theory $(\mathbb{N},+,f)$ admits QE  and thus the theory is complete and decidable \cite{Semenov84}. 
Exponential functions are compatible with addition functions, and therefore {$\paexp$} is decidable 
by \cite{Semenov84}.
Particularly, in \cite{Point86} Point gave a detailed QE algorithm for $(\mathbb{N},+,2^x)$, where $f(x)=2^x$.
Unfortunately, Point's algorithm is flawed, one problem 
 lies in the discussion to eliminate exponential terms. 
% making the algorithm fail to handle "$a_0<0$" case.
For example, after  eliminating variable $x$ in formula $\exists x. y\le 2^x \wedge x\le 10$,
the expected result should be $y\le 2^{10} =1024$,
but Point's algorithm will return a formula equivalent to $y \le 2^6=64$.
% 这里好像还有另一个与Francoise一起的作者

In the next section, within Point's framework, we give a revised algorithm
 to the decision problem of {$\paexp$} with 
both subtle improvements and crucial corrections, that 
provides a crucial step to the decision procedure of string constraints with {\parseInt} function according to 
Theorem~\ref{thm:string-parInt}.

%\section{Quantifier Elimination Algorithm for $(\mathbb{N},+,2^x)$}
Based on Point's work \cite{Point86}, we present a  revised QE algorithm for the formula of the form $Qx.\theta(x,\bar{y})$, 
where $Q$ is a quantifier and $\theta(x,\bar{y})$ is a quantifier-free formula. 
Since $\forall x. F = \neg \exists x. \neg F$, we further assume the quantifier $Q$ to be the existential quantifier, 
that is, $\exists x.\theta(x,\bar{y})$. 
For eliminating quantifiers in arbitrary formula, 
we apply the algorithm to the innermost quantified formula and repeat this procedure until all quantifiers are eliminated.

We divide the whole QE algorithm into 4 steps.
The first step $\textbf{Normalization}$ can be viewed as a pre-processing step, which substitutes ``complex" terms like $2^{2^x}$, $l_2(3x+y)$ with ``simple" terms by introducing new variables $x_i$. Now we move forward to handle a ``simpler" formula, however, at the cost of more quantified variables.

The other three steps undertake the task to eliminate the introduced variables $x_i$ and the original variable $x$ (will be denoted by $x_0$) one by one. 
The main body of \textbf{QE-with-Order} step is a loop to enumerate all possible orders among quantified variables. 
In each iteration, according the given (decreasing) order (corresponding to a for loop), we invoke \textbf{QE-exp} or 
\textbf{QE-linear} to eliminate these variables one by one. 
%depending on  eliminates the maximal variable in the remaining $\bar{x}$. So it is necessary to specify an ordering for $\bar{x}$ at the beginning, as we will see, this is the main reason for the complexity of this algorithm.
During each iteration of the inner loop (the for loop), 
if the maximal $x_i$ in $\bar{x}$ occurs in an exponential term in an atomic formula, 
we will invoke \textbf{QE-exp} to produce a formula equivalent to the atom where $x_i$ occurs linearly;  
otherwise, if $x_i$ occurs linearly in the formula, \textbf{QE-linear} will be invoked to eliminate all occurrences of $x_i$, 
and this procedure is similar to the classic QE algorithm for PA.

@@

Make substitutions by introducing new variables according to the \textit{while} statements in the pseudo-code.For consistency, we rename the original $x$ to be $x_0$ and assume $n$ new quantified variables are introduced.

After introducing new variables and substitution, 
we obtain a formula where if an exponential term contains a quantified variable $x_i(0\le i\le n)$, the term should be $2^{x_i}$.
We then collect all terms with $x_i(0\le i\le n)$ together, 
it will be of the form $s(\bar{x})\Def \sum_{i=0}^{n} a_i 2^{x_i} + \sum_{i=0}^n b_i x_i$, where $a_i,b_i(0\le i \le n)$ are all constants.
Other terms including constants and terms of $\bar{y}$ are collected,  
denoted by $t(\bar{y})$. 
Since $\bar{y}$ are free variables, 
we will regard $t(\bar{y})$ as a constant.
Inequalities and equalities will all be transformed into $s(\bar{x})\le t(\bar{y})$, for example, $s(\bar{x}) = t(\bar{y})$ will be replaced by $s(\bar{x})\le t(\bar{y})\wedge t(\bar{y})\le s(\bar{x})$.

@@

    $\theta' :=  \theta(x,\bar{y})$ \;
    Transform $\theta'$ into NNF\;
    \tcp{ i is used for counting the introduced variables }
    i = 1\; 
    \tcp{ $\bar{y}$ are regarded as constants} 
    \While{there is a term $l_2(t)$, $t$ is not a constant}
    {
        $\theta' :=  \theta'[x_i/l_2(t)] \wedge (2^{x_i}\le t) \wedge (t\le 2^{x_i+1}-1)$\;
        $i:= i+1$\;
    }
    \While{there is a term $2^t$, $t$ is not a variable $x_j(j<i)$ or a constant}
    {
        $\theta' :=  \theta'[2^{x_i}/2^t]\wedge (x_i\le t) \wedge (t \le x_i)$\;
        $i:= i+1$\;
    }
    $n:= i-1$\;
    \tcp{ Collect quantified variables $x_i$}
    Transform all atoms into forms $s(\bar{x})\le t(\bar{y})$, $k|s(\bar{x})+t(\bar{y})$ or $\neg  (k|s(\bar{x})+t(\bar{y}))$\;
    Return $\exists x_0,...,\exists x_n. \theta'$

@@

\subsection{QE-with-order} 
During the \textbf{Normalization} step, 
we simplify the origin formula at the cost of more quantified variables. 
Suppose we get $\exists \bar{x}.\theta(\bar{x},\bar{y})$,
and it has $n$ new variables $x_i$ ($x$ is denoted by $x_0$).
As a consequence, we need to eliminate all quantified variables one by one.

To the end, we denote the set of all orders among the $n+1$ variables by 
 $S_{n+1}$, and then enumerate these orders one by one in the outer \textit{for} loop. 
 For the considered order $\sigma$, we first add the ordering information to 
 the quantifier free formula, and then according to the 
 order, to eliminate $x_{\sigma(i)}$ for $i=n$ to $0$ by invoking \textbf{QE-exp} first, and then invoking \textbf{QE-linear} (the inner \textit{for} loop). 
 In each iteration of the inner \textit{for} loop, 
if $x_{\sigma{(i)}}$ occurs in an exponential term in an atomic formula, 
\textbf{QE-exp} is invoked and a formula   equivalent to the atom is returned, in which $x_i$ occurs linearly;  
then $x_i$ occurs linearly in the formula, thus \textbf{QE-linear} is further invoked to eliminate all occurrences of $x_i$. 
The returned formula is the disjunction of all formulas resulted in all iterations of the outer \textit{for} loop.

\iffalse
put , We now specify an order of $\bar{x}$, and recursively eliminate the maximal element in the remaining $\bar{x}$. 
That is to say,
every execution of \textbf{QE-exp} and \textbf{QE-linear} will eliminate the largest element $x_i$ in $\bar{x}$.
However, as you may notice, if there is no clue about the order of $\bar{x}$, 
we will have to use the idea of permutation group and the number of sub-cases will blow up, in the worst case, $(n+1)!$ cases.

Here is how \textbf{QE-exp} and \textbf{QE-linear} are invoked in
\textbf{Ordering and QE}. 
\fi

@@

        \tcp{recursively eliminate the maximal $x_i$ in $\bar{x}$}
        %\znj{How to effectively determine the maximal $x_i$?} 
        \For{i from n to 0}
        {
            \tcp{if $x_{\sigma(i)}$ occurs exponentially in $\theta_\sigma$}
            $\theta_{\sigma}:= \textbf{QE-exp}(x_{\sigma(i)},\theta_{\sigma})$\;
            \tcp{now $x_{\sigma(i)}$ occurs only linearly in $\theta_\sigma$}
            $\theta_{\sigma}:= \textbf{QE-linear}(x_{\sigma(i)},\theta_{\sigma})$\;
        }
        $\phi :=  \phi \vee \theta_{\sigma}$\;
    }
\iffalse
    \For{i from $0$ to $n$}
    {
        \tcp{ specify the maximal quantified variable}
        $\theta_i :=  \theta(\bar{x},\bar{y}) \wedge (\bigwedge_{0\le j \le n} x_i - x_j \ge 0)$\;
        \tcp{ eliminate $x_i$}
        \While{$x_i$ occurs in an exponential term in formula $\tau(x_i,\bar{x},\bar{y})$}{
            \tcp{ QE-exp outputs a formula equivalent with $\tau(x_i,\bar{x},\bar{y})$ where $x_i$ occurs linearly }
            $\theta_i :=  \theta_i[\text{QE-exp}(\tau(x_i,\bar{x},\bar{y}))/\tau(x_i,\bar{x},\bar{y})]$\;
        }   
        transform $\theta_i$ into disjunction normal form $\theta_i = \bigvee_s \theta_{i,s}$\;
        \For{ every $\theta_{i,s}$ }{
            $\theta'_{i,s} :=  \text{QE-linear}(x_i,\theta_{i,s})$
        }
        $\theta'_i =\bigvee_s \theta'_{i,s}$\;
    }
\fi
    output $\phi$
    \caption{QE-with-order}
\end{algorithm}
%\znj{I cannot understand why you need to do QE for all permutation $\sigma$? I think 
%one is enough.}

\subsection{QE-exp}

\textbf{QE-exp} and \textbf{QE-linear} are the most technical parts of our QE algorithm.
As mentioned, 
\textbf{QE-exp} takes $\theta_{\sigma}(\bar{x},\bar{y})$ and  variable $x_{\sigma(i)}$ (the maximal variable among all quantified variables that are not eliminated yet according to $\sigma$) as 
inputs, and  outputs an equivalent formula in which $x_{\sigma(i)}$ occurs linearly.
The ideal case is that the formula $\theta(\bar{x},\bar{y})$ itself contains no $2^{x_{\sigma(i)}}$ terms, so we can omit this step and directly go to \textbf{QE-linear}.
For simplicity, we will abuse   $x_i$ for $x_{\sigma(i)}$, 
and  $\bar{x}$ for $(x_{\sigma(0)},...,x_{\sigma(i-1)})$ ( remind that $x_{\sigma(i+1)},...,x_n$ have been eliminated already).

After normalization, the formula $\theta(\bar{x},\bar{y})$ contains atoms of three forms corresponding to 
the predicates $\le$, $|$ and negation of $|$, respectively.
The problem will be discussed in 2 cases depending on the form of the atom $\tau$ that contains $2^{x_i}$, 
corresponding to \textbf{QE-exp-ineq} or \textbf{QE-exp-div}.

We first give the whole procedure of \textbf{QE-exp},
and then provide the details of the two sub-routines.
\begin{algorithm}[t]
    \SetAlgoLined
    \KwIn{$x_i$ and $\theta$, $x_i$ is larger than $x_0$ to $x_{i-1}$
    and occurs exponentially in $\theta$} 
    \KwOut{a formula equivalent to $\theta$ where $x_i$ occurs linearly}

    \While{there is an atom $\tau(x_i,\bar{x},\bar{y})$ contains $2^{x_i}$}
    {
        $\Psi:= \textit{False}$\;
        \eIf{$\tau$ is of the form $a_i 2^{x_i}+\sum_{j=0}^{i-1} a_j2^{x_j} + \sum_{k=0}^{i}b_k x_k \le t(\bar{y})$}
        { 
            \tcp{ \textbf{QE-exp-ineq} case}
            $A :=  \sum_{j=0}^{i-1}|a_j|$, $B:= b_i + \sum_{j=0}^{i-1}|b_j|$\;
            $B':= 2(l_2(B)+3)$, $\delta:=  l_2(A)+3$\;
            If $a_i>0$ then $\alpha := l_2(t(y))-l_2(a_i)$
            else $\alpha := l_2(-t(y))-l_2(-a_i)$\;

@@

            $\quad \vee (x_i \ge \alpha +2 \wedge x_i\ge B' \wedge \bigvee_{0\le k \le \delta}(x_i = x_{i-1}+k \wedge \tau[x_{i-1}+k/x_i]))
            $\;
            \eIf{$a_i>0$}
            {
                $\Psi := \rho_1 \vee \rho_2 \vee \rho_3 \vee \rho_4 
                \vee [x_i \le \alpha -1 \wedge x_i \ge B' \wedge x_i \ge x_{i-1}+\delta]$
            }
            {
                $\Psi := \rho_1 \vee \rho_2 \vee \rho_3 \vee \rho_4 
                \vee [x_i \ge \alpha +2 \wedge x_i \ge B' \wedge x_i \ge x_{i-1}+\delta]$    
            }
            }
        {
            \tcp{ \textbf{QE-exp-div} case, the atom is of the form $d|t(x_i,\bar{x},\bar{y})$}
            \tcp{ let $d=2^r d_0$ where $d_0$ is odd}
            $\rho_5 := \bigvee_{p=0}^{r-1} [\tau(p,\bar{x},\bar{y})\wedge x=p]$\;
            $\rho_6 := \bigvee_{q=0}^{\phi(d_0)-1} [d|(a_i 2^{r+q}+\sum_{j=0}^{i-1} a_j 2^{x_j}
            +  \sum_{k=0}^{i} b_k x_k+t(\bar{y})) \wedge \phi(d_0)|(x_i-r-q) \wedge x_i \ge r]
            $\;
            $\Psi :=  \rho_5 \vee \rho_6$
        }
        replace $\tau$ by $\Psi$ in $\theta$\;
    }
    \caption{QE-exp}
\end{algorithm}

\subsubsection{QE-exp-ineq}

@@

\subsubsection{QE-exp-div}

If $x_i$ occurs exponentially in an atom $\tau(x_i,\bar{x},\bar{y})$ of the form
$$\tau(x_i,\bar{x},\bar{y}) \Def  d| a_i 2^{x_i}+\sum_{j=0}^{i-1} a_j 2^{x_j} + \sum_{k=0}^{i} b_k x_k+t(\bar{y}) \quad (a_i \neq 0)\,,$$
the algorithm \textbf{QE-exp-div} outputs an equivalent formula without $2^{x_i}$ terms.
The idea is that $a_i 2^x$ modulo $d$ is a periodic function when $x$ is large enough and the period can be computed.

@@

\subsection{QE-linear}

After \textbf{QE-exp}, 
we obtain a formula (still denoted by $\theta(x_i,\bar{x},\bar{y})$) without $2^{x_i}$ terms,
i.e., $x_i$ occurs only linearly.
Now we wish to construct a formula without $x_i$ equivalent to $\exists x_i.\theta(x_i,\bar{x},\bar{y})$.
This can be done by 
following Cooper's QE algorithm for PA,
treating all $x_j (j<i)$ as free variables (like $\bar{y}$).
The procedure contains 3 steps.

@@

    Collect terms of $x_i$ in each atom\;
    Let $d$ to be the least common multiple of all coefficients of $x_i$\;
    Multiply each atom by a factor so that coefficient of $x_i $ is $d$\;
    Let $\theta'(x_i'):= \theta[x_i'/dx_i]$\;
    $\theta'(x_i') := \theta'(x_i') \wedge d| x_i'$\;
    \tcp{ Now the atoms have 3 different forms}
    \tcp{ $t_l(\bar{x},\bar{y})\le x_i'$, $x_i' \le t_u(\bar{x},\bar{y})$, $k_m|x_i'+t_m(\bar{x},\bar{y})$(or its negation)}
    \tcp{ where $l\in L,u\in U,m\in M$ are the index sets}
    $\delta :=  lcm\{k_m|m\in M\}$\;
    \For{all atom $\tau_l \in L$}
    {
        $\theta'_{+\infty}:=  \theta'\{\textit{true}/\tau_l\}$
    }
    \For{all atom $\tau_u \in U$}
    {
        $\theta'_{+\infty}:=  \theta'_{+\infty}\{\textit{false}/\tau_u\}$
    }
    output $\bigvee_{j=0}^{\delta-1} \theta'_{+\infty}[j/x_i'] \vee \bigvee_{j=0}^{\delta-1} \bigvee_{u\in U}\theta'[t_u(\bar{x},\bar{y})-j/x_i']$
    \caption{QE-linear (Cooper's QE algorithm for PA)}
\end{algorithm}

@@

Point's algorithm \cite{Point86} has four steps,
corresponding to the four steps of our algorithm respectively.
So we will use the names of our steps to refer to them. 
The divisible predicate $n|x$ is not included in his theory,
instead, the division operation $\frac{x}{n}$ where $n\in \mathbb{N}$ is allowed.
The \textbf{Normalization} and \textbf{QE-with-order} steps in two algorithms are similar,
with some minor differences due to the setting of the theory.
The main differences between the two algorithms lies in the following aspects.

In \textbf{QE-exp},
we adopt the same idea to find sufficient conditions
and necessary conditions for the formula to holds.
But our strategy for choosing parameters are easier to understand and use.
A flaw lies in the $a_i<0$ case in Point's algorithm 
\footnote{
In the $a_i<0$ case of \textbf{QE-exp-ineq},
Point's algorithm assumes falsely $t(y)\ge 0$.
The reason for this mistake might be the confusion of 
two subtraction operators.
The example mentioned in \ref{sec:3.2} shows this case.
},
our \textbf{QE-exp-ineq} corrects this flaw by 
treating $a_i<0$ case similarly to the $a_i>0$ case.

>>>>

\begin{definition}
    For a strictly increasing function $f:\mathbb{N}\mapsto \mathbb{N}$, 
    $f$ is said to be \emph{compatible with addition}
    if for every $m\in \mathbb{N}^+$,  $f$ modulo $m$ is periodic, 
    and for any term $A(x)=\sum_{1\le i\le n} a_i f(x+b_i)$, 
    where $n \in \mathbb{N}^+, a_i ,b_i \in \mathbb{Z}$,
    one of the following holds:
    \begin{itemize}
        \item $A(x)$ is bounded; 
        \item there exists a constant $\Delta_A\in \mathbb{N}^+$ such that 
        $\forall x. A(x+\Delta_A)\ge f(x)$;
        \item there exists a constant $\Delta_A\in \mathbb{N}^+$ such that 
        $\forall x. -A(x+\Delta_A)\ge f(x)$.
    \end{itemize}
\end{definition}

Semenov proved that for any function $f$ that is compatible with addition, the first order theory $(\mathbb{N},+,f)$ admits QE  and thus is complete and decidable \cite{Semenov84}. Exponential functions are a typical class of compatible with addition functions, and therefore {$\paexp$} is decidable. Particularly, in \cite{Point86} 
Point gave a detailed QE algorithm for $(\mathbb{N},+,2^x)$.
%下面这些感觉可以不用说，或者放在后面
%Unfortunately, Point's algorithm is flawed, one problem lies in the discussion to eliminate exponential terms. For example, after  eliminating variable $x$ in formula $\exists x. y\le 2^x \wedge x\le 10$, the expected result should be $y\le 2^{10} =1024$, but Point's algorithm will return a formula equivalent to $y \le 2^6=64$.

In this section, within Point's framework, we give a revised algorithm to the decision problem of {$\paexp$} with both improvements and corrections, that provides a crucial step to the decision procedure of string constraints with {\parseInt} function according to Theorem~\ref{thm:string-parInt}.

%\section{Quantifier Elimination Algorithm for $(\mathbb{N},+,2^x)$}
% Based on Point's work \cite{Point86}, 
We will present a revised QE algorithm for {$\paexp$} formulae of the form $Qx.\theta(x,\bar{y})$, where $Q$ is a quantifier and $\theta(x,\bar{y})$ is a quantifier-free formula. Since $\forall x. F = \neg \exists x. \neg F$, we further assume the quantifier is an existential quantifier, that is, $\exists x.\theta(x,\bar{y})$. For eliminating quantifiers in arbitrary {$\paexp$} formula, we need to apply the algorithm to the innermost quantified formula and repeat this procedure until all quantifiers are eliminated.

The whole QE algorithm can be divided into 4 steps. The first step \textbf{Normalization} can be viewed as a pre-processing step, which substitutes ``complex" terms like $2^{2^x}$, $l_2(3x+y)$ with ``simple" terms by introducing new variables. After \textbf{Normalization}, we move forward to handle a ``simpler" formula, however, at the cost of more quantified variables.

The other three steps undertake the task to eliminate the newly introduced variables in \textbf{Normalization} step, denoted by $\{x_i:1\le i\le n\}$, and the original variable $x$ (will be denoted by $x_0$) one by one. 
The main body of \textbf{Enumerate-Orders} step is a loop to enumerate all possible orders among quantified variables $\{x_i : 0\le i \le n\}$. 
In each iteration, according the given (decreasing) order of quantified variables (corresponding to an outer \emph{for} loop), we invoke \textbf{Elim-exp} and \textbf{Elim-linear} to eliminate these variables one by one.

During each iteration of the inner \emph{for} loop, if the maximal $x_i$ in $\bar{x}$ occurs in an exponential term in an atomic formula $\tau$, we will invoke \textbf{Elim-exp} to produce a new formula where $x_i$ occurs linearly equivalent to $\tau$. Then \textbf{Elim-linear} will eliminate all linear occurrences of $x_i$, which is similar to the classic Cooper's QE algorithm for PA \cite{Cooper72}.

@@

Make substitutions by introducing new variables according to the \textit{while} statements in the pseudo-code. For consistency, we rename the original $x$ to be $x_0$ and assume $n$ new quantified variables are introduced.

%After introducing new variables and substitution, 
After substitutions, we obtain a formula where if an exponential term contains a quantified variable $x_i$, the term must be $2^{x_i}$.
We then collect all terms with $x_i (0\le i\le n)$ together, it will be of the form $s(\bar{x}) \Def \sum_{i=0}^{n} a_i 2^{x_i} + \sum_{i=0}^n b_i x_i$, where $a_i,b_i (0\le i \le n)$ are all constants.
Other terms including constants and terms of free variables $\bar{y}$ are collected, denoted by $t(\bar{y})$. Since $\bar{y}$ are free variables, we will regard $t(\bar{y})$ as a constant. Inequalities and equalities will all be transformed into $s(\bar{x})\le t(\bar{y})$, for example, $s(\bar{x}) = t(\bar{y})$ will be replaced by $s(\bar{x})\le t(\bar{y})\wedge t(\bar{y})\le s(\bar{x})$.

@@

    $\theta' :=  \theta(x,\bar{y})$ \;
    Transform $\theta'$ into NNF\;
    \tcp{ i is used for counting the introduced variables }
    $i := 1$\; 
    \tcp{ $\bar{y}$ are regarded as constants} 
    \While{there is a term $l_2(t)$, $t$ is not a constant}
    {
        $\theta' :=  \theta'[x_i/l_2(t)] \wedge (2^{x_i}\le t) \wedge (t\le 2^{x_i+1}-1)$\;
        $i:= i+1$\;
    }
    \While{there is a term $2^t$, $t$ is not a variable $x_j(j<i)$ or a constant}
    {
        $\theta' :=  \theta'[2^{x_i}/2^t]\wedge (x_i\le t) \wedge (t \le x_i)$\;
        $i:= i+1$\;
    }
    $n:= i-1$\;
    \tcp{ Collect quantified variables}
    Transform all atoms of $\theta'$ into forms $s(\bar{x})\le t(\bar{y})$, $k|s(\bar{x})+t(\bar{y})$ or $\neg  (k|s(\bar{x})+t(\bar{y}))$\;
    Return $\exists x_0,...,\exists x_n. \theta'$

@@

\subsection{Enumerate-Orders} 
In the \textbf{Normalization} step, we simplify the origin formula at the cost of more quantified variables. 
Suppose we get $\exists \bar{x}.\theta(\bar{x},\bar{y})$,
and it has $n$ new variables $x_i$ ($x$ is denoted by $x_0$).
As a consequence, we need to eliminate all quantified variables one by one.

To the end, we denote the set of all orders among the $n+1$ variables by 
 $S_{n+1}$, and then enumerate these orders one by one in the outer \textit{for} loop. 
 For the considered order $\sigma$, we first add the ordering information to 
 the quantifier free formula, and then according to the 
 order, to eliminate $x_{\sigma(i)}$ for $i=n$ to $0$ by invoking \textbf{Elim-exp} first, and then invoking \textbf{Elim-linear} (the inner \textit{for} loop). 
 In each iteration of the inner \textit{for} loop, 
if $x_{\sigma{(i)}}$ occurs in an exponential term in an atomic formula, 
\textbf{Elim-exp} is invoked and a formula   equivalent to the atom is returned, in which $x_i$ occurs linearly;  
then $x_i$ occurs linearly in the formula, thus \textbf{Elim-linear} is further invoked to eliminate all occurrences of $x_i$. 
The returned formula is the disjunction of all formulas resulted in all iterations of the outer \textit{for} loop.

\iffalse
put , We now specify an order of $\bar{x}$, and recursively eliminate the maximal element in the remaining $\bar{x}$. 
That is to say,
every execution of \textbf{Elim-exp} and \textbf{Elim-linear} will eliminate the largest element $x_i$ in $\bar{x}$.
However, as you may notice, if there is no clue about the order of $\bar{x}$, 
we will have to use the idea of permutation group and the number of sub-cases will blow up, in the worst case, $(n+1)!$ cases.

Here is how \textbf{Elim-exp} and \textbf{Elim-linear} are invoked in
\textbf{Ordering and QE}. 
\fi

@@

        \tcp{recursively eliminate the maximal $x_i$ in $\bar{x}$}
        %\znj{How to effectively determine the maximal $x_i$?} 
        \For{i from n to 0}
        {
            \tcp{if $x_{\sigma(i)}$ occurs exponentially in $\theta_\sigma$}
            $\theta_{\sigma}:= \textbf{Elim-exp}(x_{\sigma(i)},\theta_{\sigma})$\;
            \tcp{now $x_{\sigma(i)}$ occurs only linearly in $\theta_\sigma$}
            $\theta_{\sigma}:= \textbf{Elim-linear}(x_{\sigma(i)},\theta_{\sigma})$\;
        }
        $\phi :=  \phi \vee \theta_{\sigma}$\;
    }
\iffalse
    \For{i from $0$ to $n$}
    {
        \tcp{ specify the maximal quantified variable}
        $\theta_i :=  \theta(\bar{x},\bar{y}) \wedge (\bigwedge_{0\le j \le n} x_i - x_j \ge 0)$\;
        \tcp{ eliminate $x_i$}
        \While{$x_i$ occurs in an exponential term in formula $\tau(x_i,\bar{x},\bar{y})$}{
            \tcp{ Elim-exp outputs a formula equivalent with $\tau(x_i,\bar{x},\bar{y})$ where $x_i$ occurs linearly }
            $\theta_i :=  \theta_i[\text{Elim-exp}(\tau(x_i,\bar{x},\bar{y}))/\tau(x_i,\bar{x},\bar{y})]$\;
        }   
        transform $\theta_i$ into disjunction normal form $\theta_i = \bigvee_s \theta_{i,s}$\;
        \For{ every $\theta_{i,s}$ }{
            $\theta'_{i,s} :=  \text{Elim-linear}(x_i,\theta_{i,s})$
        }
        $\theta'_i =\bigvee_s \theta'_{i,s}$\;
    }
\fi
    output $\phi$
    \caption{Enumerate-Orders}
\end{algorithm}
%\znj{I cannot understand why you need to do QE for all permutation $\sigma$? I think 
%one is enough.}

\subsection{Elim-exp}

\textbf{Elim-exp} and \textbf{Elim-linear} are the most technical parts of our QE algorithm.
As mentioned, 
\textbf{Elim-exp} takes $\theta_{\sigma}(\bar{x},\bar{y})$ and  variable $x_{\sigma(i)}$ (the maximal variable among all quantified variables that are not eliminated yet according to $\sigma$) as 
inputs, and  outputs an equivalent formula in which $x_{\sigma(i)}$ occurs linearly.
The ideal case is that the formula $\theta(\bar{x},\bar{y})$ itself contains no $2^{x_{\sigma(i)}}$ terms, so we can omit this step and directly go to \textbf{Elim-linear}.
For simplicity, we will abuse   $x_i$ for $x_{\sigma(i)}$, 
and  $\bar{x}$ for $(x_{\sigma(0)},...,x_{\sigma(i-1)})$ ( remind that $x_{\sigma(i+1)},...,x_n$ have been eliminated already).

After normalization, the formula $\theta(\bar{x},\bar{y})$ contains atoms of three forms corresponding to 
the predicates $\le$, $|$ and negation of $|$, respectively.
The problem will be discussed in 2 cases depending on the form of the atom $\tau$ that contains $2^{x_i}$, 
corresponding to \textbf{Elim-exp-ineq} or \textbf{Elim-exp-div}.

We first give the whole procedure of \textbf{Elim-exp},
and then provide the details of the two sub-routines.
\begin{algorithm}[t]
    \SetAlgoLined
    \KwIn{$x_i$ and $\theta$, $x_i$ is larger than $x_0$ to $x_{i-1}$
    and occurs exponentially in $\theta$} 
    \KwOut{a formula equivalent to $\theta$ where $x_i$ occurs linearly}

    \While{there is an atom $\tau(x_i,\bar{x},\bar{y})$ contains $2^{x_i}$}
    {
        $\Psi:= \textit{False}$\;
        \eIf{$\tau$ is of the form $a_i 2^{x_i}+\sum_{j=0}^{i-1} a_j2^{x_j} + \sum_{k=0}^{i}b_k x_k \le t(\bar{y})$}
        { 
            \tcp{ \textbf{Elim-exp-ineq} case}
            $A :=  \sum_{j=0}^{i-1}|a_j|$, $B:= b_i + \sum_{j=0}^{i-1}|b_j|$\;
            $B':= 2(l_2(B)+3)$, $\delta:=  l_2(A)+3$\;
            If $a_i>0$ then $\alpha := l_2(t(y))-l_2(a_i)$
            else $\alpha := l_2(-t(y))-l_2(-a_i)$\;

@@

            $\quad \vee (x_i \ge \alpha +2 \wedge x_i\ge B' \wedge \bigvee_{0\le k \le \delta}(x_i = x_{i-1}+k \wedge \tau[x_{i-1}+k/x_i]))
            $\;
            \eIf{$a_i>0$}
            {
                $\Psi := \rho_1 \vee \rho_2 \vee \rho_3 \vee \rho_4 
                \vee [x_i \le \alpha -1 \wedge x_i \ge B' \wedge x_i \ge x_{i-1}+\delta]$
            }
            {
                $\Psi := \rho_1 \vee \rho_2 \vee \rho_3 \vee \rho_4 
                \vee [x_i \ge \alpha +2 \wedge x_i \ge B' \wedge x_i \ge x_{i-1}+\delta]$    
            }
            }
        {
            \tcp{ \textbf{Elim-exp-div} case, the atom is of the form $d|t(x_i,\bar{x},\bar{y})$}
            \tcp{ let $d=2^r d_0$ where $d_0$ is odd}
            $\rho_5 := \bigvee_{p=0}^{r-1} [\tau(p,\bar{x},\bar{y})\wedge x=p]$\;
            $\rho_6 := \bigvee_{q=0}^{\phi(d_0)-1} [d|(a_i 2^{r+q}+\sum_{j=0}^{i-1} a_j 2^{x_j}
            +  \sum_{k=0}^{i} b_k x_k+t(\bar{y})) \wedge \phi(d_0)|(x_i-r-q) \wedge x_i \ge r]
            $\;
            $\Psi :=  \rho_5 \vee \rho_6$
        }
        replace $\tau$ by $\Psi$ in $\theta$\;
    }
    \caption{Elim-exp}
\end{algorithm}

\subsubsection{Elim-exp-ineq}

@@

\subsubsection{Elim-exp-div}

If $x_i$ occurs exponentially in an atom $\tau(x_i,\bar{x},\bar{y})$ of the form
$$\tau(x_i,\bar{x},\bar{y}) \Def  d| a_i 2^{x_i}+\sum_{j=0}^{i-1} a_j 2^{x_j} + \sum_{k=0}^{i} b_k x_k+t(\bar{y}) \quad (a_i \neq 0)\,,$$
the algorithm \textbf{Elim-exp-div} outputs an equivalent formula without $2^{x_i}$ terms.
The idea is that $a_i 2^x$ modulo $d$ is a periodic function when $x$ is large enough and the period can be computed.

@@

\subsection{Elim-linear}

After \textbf{Elim-exp}, 
we obtain a formula (still denoted by $\theta(x_i,\bar{x},\bar{y})$) without $2^{x_i}$ terms,
i.e., $x_i$ occurs only linearly.
Now we wish to construct a formula without $x_i$ equivalent to $\exists x_i.\theta(x_i,\bar{x},\bar{y})$.
This can be done by 
following Cooper's QE algorithm for PA,
treating all $x_j (j<i)$ as free variables (like $\bar{y}$).
The procedure contains 3 steps.

@@

    Collect terms of $x_i$ in each atom\;
    Let $d$ to be the least common multiple of all coefficients of $x_i$\;
    Multiply each atom by a factor so that coefficient of $x_i $ is $d$\;
    Let $\theta'(x_i'):= \theta[x_i'/dx_i]$\;
    $\theta'(x_i') := \theta'(x_i') \wedge d| x_i'$\;
    \tcp{ Now the atoms have 3 different forms}
    \tcp{ $t_l(\bar{x},\bar{y})\le x_i'$, $x_i' \le t_u(\bar{x},\bar{y})$, $k_m|x_i'+t_m(\bar{x},\bar{y})$(or its negation)}
    \tcp{ where $l\in L,u\in U,m\in M$ are the index sets}
    $\delta :=  lcm\{k_m|m\in M\}$\;
    \For{all atom $\tau_l \in L$}
    {
        $\theta'_{+\infty}:=  \theta'\{\textit{true}/\tau_l\}$
    }
    \For{all atom $\tau_u \in U$}
    {
        $\theta'_{+\infty}:=  \theta'_{+\infty}\{\textit{false}/\tau_u\}$
    }
    output $\bigvee_{j=0}^{\delta-1} \theta'_{+\infty}[j/x_i'] \vee \bigvee_{j=0}^{\delta-1} \bigvee_{u\in U}\theta'[t_u(\bar{x},\bar{y})-j/x_i']$
    \caption{Elim-linear (Cooper's QE algorithm for PA)}
\end{algorithm}

@@

Point's algorithm \cite{Point86} has four steps,
corresponding to the four steps of our algorithm respectively.
So we will use the names of our steps to refer to them. 
The divisible predicate $n|x$ is not included in his theory,
instead, the division operation $\frac{x}{n}$ where $n\in \mathbb{N}$ is allowed.
The \textbf{Normalization} and \textbf{Enumerate-Orders} steps in two algorithms are similar,
with some minor differences due to the setting of the theory.
The main differences between the two algorithms lies in the following aspects.

In \textbf{Elim-exp},
we adopt the same idea to find sufficient conditions
and necessary conditions for the formula to holds.
But our strategy for choosing parameters are easier to understand and use.
A flaw lies in the $a_i<0$ case in Point's algorithm 
\footnote{
In the $a_i<0$ case of \textbf{Elim-exp-ineq},
Point's algorithm assumes falsely $t(y)\ge 0$.
The reason for this mistake might be the confusion of 
two subtraction operators.
The example mentioned in \ref{sec:3.2} shows this case.
},
our \textbf{Elim-exp-ineq} corrects this flaw by 
treating $a_i<0$ case similarly to the $a_i>0$ case.
